Analysis of the inference of ratings and rankings on Higher Order Networks with complex topologies

### 2025-09-08 draft/draft.tex
progress

Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$.
For each distinct pair of teams $i$ and $j$, assume that a pairwise comparison score $f_{ij}=-f_{ji}$ is available.
Intuitively, $f_{ij}>0$ indicates that team $i$ should be rated higher than team $j$, whereas $f_{ij}<0$ suggests the opposite.

A natural approach to inferring ratings consistent with these comparisons is to assign scores $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies:
\begin{equation}
\label{eq:1}
\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
\end{equation}
Once the team ratings $w_0,w_1,\dots,w_n$ have been inferred, the ranking $r_0,r_1,\dots,r_n$ is obtained by requiring that $r_i<r_j$ whenever $w_i>w_j$.
This setup leads to a well-posed optimization problem that admits an elegant reformulation in terms of cochains over simplicial complexes.
The resulting framework for rating and ranking inference is known as \emph{HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.

@@

To review the HodgeRank method, we summarize the basic elements of discrete exterior calculus on simplicial complexes.  
Let $V=\{0,1,\dots,n\}$ denote a set of vertices.  
A simplicial complex is a collection
\[
K = K_0 \cup K_1 \cup \dots \cup K_n
\]
where
\[
K_k \subseteq \bigl\{\{i_0,\dots,i_k\} : 0\leq i_0<\dots<i_k\leq n,\ i_j\in V\bigr\},
\]
of $k$-simplices, i.e. unordered subsets $\{i_0,\dots,i_k\}$ of the vertex set satisfying the closure condition: if $s\in K$ and $r\subset s$ with $r\neq \emptyset$, then $r\in K$.

The set $K$ of simplices is associated with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.
We equip $C$ with the trivial inner product
\[
e_r \cdot e_s = \delta_{rs}, \qquad r,s \in K.
\]
Using Einstein summation convention, any chain $c\in C$ can be written as $c=c^se_s$ with coefficients $c^s\in \mathbb{R}$.
The chain space decomposes into
\[
C = C_0 \oplus C_1 \oplus C_2,
\]
where $C_k$ is the subspace spanned by $\{e_s:s\in K_k\}$.

Dually, the space of cochains $C^*$ is spanned by $\{e^s:s\in K\}$, where $e^r(e_s)=\delta^r{}_s$.
Any $f\in C^*$ decomposes as $f=f_se^s$ with coefficients $f_s\in \mathbb{R}$, and
\[
C^* = C_0^* \oplus C_1^* \oplus C_2^*.
\]

By Riesz’s theorem, every cochain $f\in C^*$ corresponds uniquely to a chain $f^\flat \in C$ such that $f(c) = f^\flat \cdot c$ for all $c\in C$.
This induces an inner product on $C^*$:
\[
f\cdot g = f^\flat \cdot g^\flat.
\]
The map $f\mapsto f^\flat$ is an isomorphism, with inverse denoted by $c\mapsto c^\sharp$.

@@

\noindent\textbf{Orientation and boundary.}  
An orientation is imposed on chains by requiring
\[
e_s = e_{i_0\dots i_k} = (-1)^p e_{j_0\dots j_k},
\]
for each simplex $s=\{i_0,\dots,i_k\}\in K$, where $0\leq i_0<\dots<i_k\leq n$ and $p$ is the parity of the permutation taking $(i_0,\dots,i_k)$ into $(j_0,\dots,j_k)$.

The boundary operator $\partial:C\to C$ is defined on basis elements as
\[
\partial e_{i_0\dots i_k}
=
\sum_{j=0}^k (-1)^j e_{i_0\dots i_{j-1}i_{j+1}\dots i_k}.
\]
Thus $\partial(C_k)\subseteq C_{k-1}$, with $C_{-1}=\{0\}$.
This yields the decomposition
\[
\partial = \partial_0 \oplus \partial_1 \oplus \partial_2,
\]
where $\partial_k(C_k)\subseteq C_{k-1}$ and $\partial_0=0$.

@@

\noindent\textbf{Coboundary.}  
The coboundary operator $d:C^*\to C^*$ is defined by
\[
df(c) = f(\partial c),
\]
for all $f\in C^*$ and $c\in C$.
Thus $d(C_k^*) \subseteq C_{k+1}^*$, with $C_3^*=\{0\}$, and
\[
d = d_0 \oplus d_1 \oplus d_2, \qquad d_2=0.
\]

@@

\noindent\textbf{Adjoint operators.}  
Since the inner products on $C$ and $C^*$ are bilinear, the adjoint operators $\partial^*:C\to C$ and $d^*:C^*\to C^*$ are defined by
\[
\partial^*c \cdot q = c\cdot \partial q, 
\qquad 
d^*f \cdot g = f\cdot dg,
\]
for all $c,q\in C$ and $f,g\in C^*$.
These satisfy
\[
\partial^* = \partial_0^* \oplus \partial_1^* \oplus \partial_2^*, 
\qquad 
d^* = d_0^* \oplus d_1^* \oplus d_2^*.
\]

\noindent\textbf{Hodge Laplacians.}  
The boundary and coboundary operators define the Hodge Laplacians
\[
\mathcal{L} = (\partial+\partial^*)^2 = \partial\partial^* + \partial^*\partial : C\to C,
\]
\[
L = (d+d^*)^2 = dd^* + d^*d : C^*\to C^*,
\]
where $\partial^2=(\partial^*)^2=d^2=(d^*)^2=0$.
These decompose as
\[
\mathcal{L} = \mathcal{L}_0 \oplus \mathcal{L}_1 \oplus \mathcal{L}_2, 
\qquad 
L = L_0 \oplus L_1 \oplus L_2,
\]
with
\[
\mathcal{L}_k = \mathcal{L}_k^{\downarrow}\oplus \mathcal{L}_k^{\uparrow}, 
\qquad 
L_k = L_k^{\downarrow}\oplus L_k^{\uparrow},
\]
\[
\mathcal{L}_k^{\downarrow} = \partial_k^*\partial_k, \quad 
\mathcal{L}_k^{\uparrow} = \partial_{k+1}\partial_{k+1}^*, 
\]
\[
L_k^{\uparrow} = d_k^*d_k, \quad 
L_k^{\downarrow} = d_{k-1}d_{k-1}^*.
\]

@@

\subsection{HodgeRank}

Expression~\eqref{eq:1} can be reformulated in the language of discrete calculus as
\begin{equation}
\label{eq:4}
\lVert f - d_0(w) \rVert,
\end{equation}
where the scalar components $f_{ij}$ of the $1$-cochain $f=f_{ij}e^{ij}$ encode the pairwise comparison scores, and the components $w_i$ of the $0$-cochain $w=w_ie^i$ represent the ratings.
The minimizer of~\eqref{eq:4} coincides with the solution of Eq.~\eqref{eq:3} for $k=1$.

In this setting, the set of $0$-simplices $K_0$ corresponds to the teams $V$, the set of $1$-simplices $K_1$ to the comparisons $E$, and the set of $2$-simplices $K_2$ to the closed triples of comparisons:
\[
K_2 = \bigl\{ \{i,j,k\} : \{i,j\},\{j,k\},\{i,k\}\in E \bigr\}.
\]
For $k>2$, $K_k=\emptyset$.

\section{
Methods
\label{sec:methods}
}

The dataset covers five soccer leagues: England, France, Germany, Italy and Spain.
Each league consists of $N=20$ teams, except for Germany, which has $N=18$.

in the major European soccer leagues (e.g., Premier League, La Liga, Serie A, Bundesliga, Ligue 1), the final standings are determined by a points system based on the results of matches during the season.
In the standard system, a win adds 3 points, a draw 1 point and a loss 0 points.
Each team plays every other team in the league twice (home and away) in what is called a round-robin format. 
Over the course of the season, the $i$-th team accumulates a total of $\pi_i$ points.
At the end of the season, the teams are ranked in descending order of total points.
If two or more teams have the same number of points, tiebreakers are applied. The exact rules vary slightly by league, but usually the order is something like:
Goal difference (GD): goals scored minus goals conceded.
Goals scored (GF): total number of goals scored during the season.
Head-to-head record: results of the matches between the tied teams (some leagues use this before GD).
Playoff or extra match: in rare cases (e.g., for the title, relegation, or European qualification) if all other criteria fail.
So the final standing (the ranking table at the end of the season) is determined by total points, with tiebreakers applied as needed.

\subsection{
Data
\label{sec:data}
}

For each pair $\{i,j\}$ of teams in a given league, several pairwise metrics $m_{ij}$ and $m_{ji}$ are provided.
Each metric $m_{ij}$ takes a positive value when it presumably favors the home team $i$ over away team $j$, and takes a negative value otherwise.
The different types of empirically determined metrics being considered are:
\begin{enumerate}
\item {\bf crossings:}
\item {\bf counterattacks:}
\item {\bf pressure loss:}
\item {\bf build-up time:}
\item {\bf direct play:}
\item {\bf pressure points:}
\item {\bf shots:}
\item {\bf flow rate:}
\item {\bf maintenance time:}
\item {\bf middle-zone time:} bla bla.
\end{enumerate}
Moreover, the Singular Value Decomposition of previous emprically determined metrics are considered in order to define the so-called {\em styles} which are also some kind of metric~\cite{chacoma2025data}.

\subsection{Inference of ratings and rankings}

With each metric, the coefficients
$$
f_{ij}
=
\frac{1}{2}(m_{ij}+m_{ji})
=
- f_{ji}
$$
are defined for each $i$ and $j$ such that $i<j$.
Moreover, in order to test the HodgeRank inference method, two types of ground-truth coefficients are also defined.
Namely, the rank-based coefficients $f_{ij}=i-j$ and the tournament points based coefficients $f_{ij}=\pi_j-\pi_i$.
In all cases there is an underlying fully connected network or graph with $N$ nodes, that is promoted to a simplicial complex composed of 0-simplices labeled by $i$, 1-simplices labeled by $ij$ with $i<j$ and 2-simplices labeled by $ijk$ with $i<j<k$.
Correspondingly, the coefficients $f_{ij}$ are promoted to be the components of a 1-cochain  $f$, where the anti-symmetric relation $f_{ji}=-f_{ij}$ must hold by definition.

%From a score $\phi$, a corresponding cochain $f\in C_1^*$ of coordinates $f_{ij}=\phi_{ji}-\phi_{ij}$ is defined.
%Also, the cochain $f'\in C_1^*$ of coordinates $f'_{ij}=\phi_j-\phi_i$ with team strengths $\phi_i=\sum_j \phi_{ij}$ is considered.

{\color{red}[JUAN: Write here the inference problem.]}

>>>>

Consider a set of $N=n+1$ teams labeled $\{0,1,\dots,n\}$.
For each distinct pair of teams $i$ and $j$, assume a pairwise comparison score $f_{ij}=-f_{ji}$ is available.
Informally, $f_{ij}>0$ indicates that team $i$ is rated higher than team $j$, whereas $f_{ij}<0$ suggests the opposite.

A natural approach to inferring consistent ratings is to assign scores $w_i$ ($i=0,1,\dots,n$) that minimize the total squared discrepancy
\begin{equation}
\label{eq:1}
\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
\end{equation}
Once the team ratings $w_0,w_1,\dots,w_n$ have been obtained, the ranking $r_0,r_1,\dots,r_n$ follows from the ordering condition $r_i<r_j$ whenever $w_i>w_j$.
This least-squares formulation defines a well-posed optimization problem that admits an elegant reformulation in terms of cochains over simplicial complexes.
The resulting framework for rating and ranking inference is known as \emph{HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.

@@

To review the HodgeRank method, we recall the basic elements of discrete exterior calculus on simplicial complexes.  
Let $V=\{0,1,\dots,n\}$ denote a set of vertices, and define a simplicial complex as a collection
\[
K = K_0 \cup K_1 \cup \dots \cup K_n,
\]
where
\[
K_k
\subseteq
\bigl\{
\{i_0,\dots,i_k\}
:
0\leq i_0<\dots<i_k\leq n
\bigr\}
\]
is the set of $k$-simplices, i.e., unordered subsets $\{i_0,\dots,i_k\}$ of vertices satisfying the closure condition: if $s\in K$ and $r\subset s$ with $r\neq \emptyset$, then $r\in K$.

The set $K$ of simplices is associated with a vector space $C$ of \emph{chains}, with canonical basis $\{e_s:s\in K\}$.
We equip $C$ with the inner product
\[
e_r \cdot e_s = \delta_{rs}, \qquad r,s \in K.
\]
With Einstein summation convention, any chain $c\in C$ can be written as $c=c^s e_s$ with coefficients $c^s\in \mathbb{R}$.
The chain space decomposes as
\[
C = C_0 \oplus C_1 \oplus C_2,
\]
where $C_k$ is spanned by $\{e_s:s\in K_k\}$.

The dual space $C^*$ of \emph{cochains} is spanned by $\{e^s:s\in K\}$, where $e^r(e_s)=\delta^r{}_s$.
Any $f\in C^*$ decomposes as $f=f_s e^s$ with coefficients $f_s\in \mathbb{R}$, and
\[
C^* = C_0^* \oplus C_1^* \oplus C_2^*.
\]

By Riesz’s theorem, every cochain $f\in C^*$ corresponds uniquely to a chain $f^\flat \in C$ such that $f(c) = f^\flat \cdot c$ for all $c\in C$.
This induces an inner product on $C^*$:
\[
f\cdot g = f^\flat \cdot g^\flat.
\]
The map $f\mapsto f^\flat$ is an isomorphism, with inverse denoted $c\mapsto c^\sharp$.

@@

\noindent\textbf{Orientation and boundary.}  
An orientation is fixed by requiring
\[
e_{i_0\dots i_k} = (-1)^p e_{j_0\dots j_k},
\]
for each simplex $s=\{i_0,\dots,i_k\}\in K$, where $p$ is the parity of the permutation mapping $(i_0,\dots,i_k)$ to $(j_0,\dots,j_k)$.

The boundary operator $\partial:C\to C$ is defined on basis elements as
\[
\partial e_{i_0\dots i_k}
=
\sum_{j=0}^k (-1)^j e_{i_0\dots i_{j-1}i_{j+1}\dots i_k}.
\]
Thus $\partial(C_k)\subseteq C_{k-1}$, with $C_{-1}=\{0\}$, yielding the decomposition
\[
\partial = \partial_0 \oplus \partial_1 \oplus \partial_2,
\qquad \partial_k(C_k)\subseteq C_{k-1}, \ \partial_0=0.
\]

@@

\noindent\textbf{Coboundary.}  
The coboundary operator $d:C^*\to C^*$ is defined by
\[
df(c) = f(\partial c), \qquad f\in C^*, \ c\in C.
\]
Hence $d(C_k^*) \subseteq C_{k+1}^*$, with $C_3^*=\{0\}$, and
\[
d = d_0 \oplus d_1 \oplus d_2, \qquad d_2=0.
\]

@@

\noindent\textbf{Adjoint operators.}  
The adjoints $\partial^*:C\to C$ and $d^*:C^*\to C^*$ are defined by
\[
\partial^*c \cdot q = c\cdot \partial q, 
\qquad 
d^*f \cdot g = f\cdot dg,
\]
for all $c,q\in C$ and $f,g\in C^*$.
They decompose as
\[
\partial^* = \partial_0^* \oplus \partial_1^* \oplus \partial_2^*, 
\qquad 
d^* = d_0^* \oplus d_1^* \oplus d_2^*.
\]

\noindent\textbf{Hodge Laplacians.}  
The boundary and coboundary operators define the Hodge Laplacians
\begin{eqnarray}
\mathcal{L} 
&=&
(\partial+\partial^*)^2 
=
\partial \partial^* + \partial^* \partial : C\to C,
\mbox{and}
\notag
\\
L &=& (d+d^*)^2 = dd^* + d^*d : C^*\to C^*,
\notag
\end{eqnarray}
with $\partial^2=(\partial^*)^2=d^2=(d^*)^2=0$.
These decompose as
\[
\mathcal{L} = \mathcal{L}_0 \oplus \mathcal{L}_1 \oplus \mathcal{L}_2, 
\qquad 
L = L_0 \oplus L_1 \oplus L_2,
\]
with
\[
\mathcal{L}_k = \mathcal{L}_k^{\downarrow}\oplus \mathcal{L}_k^{\uparrow}, 
\qquad 
L_k = L_k^{\downarrow}\oplus L_k^{\uparrow},
\]
\[
\mathcal{L}_k^{\downarrow} = \partial_k^*\partial_k, \quad 
\mathcal{L}_k^{\uparrow} = \partial_{k+1}\partial_{k+1}^*, 
\]
\[
L_k^{\uparrow} = d_k^*d_k, \quad 
L_k^{\downarrow} = d_{k-1}d_{k-1}^*.
\]

@@

%\subsection{HodgeRank}

The least-squares problem~\eqref{eq:1} admits a reformulation in the language of discrete calculus:
\begin{equation}
\label{eq:4}
\lVert f - d_0(w) \rVert,
\end{equation}
where the scalar entries $f_{ij}$ of the $1$-cochain $f=f_{ij}e^{ij}$ encode the pairwise comparison scores, and the components $w_i$ of the $0$-cochain $w=w_ie^i$ represent the team ratings.
The minimizer of~\eqref{eq:4} coincides with the solution of Eq.~\eqref{eq:3} for $k=1$.

In this setting, the set of $0$-simplices $K_0$ corresponds to the teams $V$, the set of $1$-simplices $K_1$ to the pairwise comparisons $E$, and the set of $2$-simplices $K_2$ to the closed triples of comparisons:
\[
K_2 = \bigl\{ \{i,j,k\} : \{i,j\},\{j,k\},\{i,k\}\in E \bigr\}.
\]
For $k>2$, $K_k=\emptyset$.

\section{Methods}
\label{sec:methods}

The dataset covers five major soccer leagues: England, France, Germany, Italy, and Spain.
Each league consists of $N=20$ teams, except Germany, which has $N=18$.

In these competitions (e.g., Premier League, La Liga, Serie A, Bundesliga, Ligue 1), the final standings are determined by a points system based on match results throughout the season.
In the standard system, a win yields 3 points, a draw 1 point, and a loss 0 points.
Each team plays every other team twice (home and away) in a round-robin format. 
Over the season, the $i$-th team accumulates $\pi_i$ total points.
At the end of the season, teams are ranked in descending order of $\pi_i$.
If two or more teams tie on points, tiebreakers are applied, typically in the following order (with minor variations across leagues):  
goal difference (GD: goals scored minus goals conceded),  
goals scored (GF),  
head-to-head record,  
and, in rare cases, a playoff or extra match.  
Thus, the final table is determined primarily by total points, with tiebreakers resolving ties.

%\subsection{Data}
\label{sec:data}

For each pair $\{i,j\}$ of teams in a given league, several pairwise metrics $m_{ij}$ and $m_{ji}$ are recorded.
Each metric $m_{ij}$ takes a positive value when it favors team $i$ over $j$, and a negative value otherwise.
The metrics considered include:
\begin{enumerate}
\item \textbf{crossings}
\item \textbf{counterattacks}
\item \textbf{pressure loss}
\item \textbf{build-up time}
\item \textbf{direct play}
\item \textbf{pressure points}
\item \textbf{shots}
\item \textbf{flow rate}
\item \textbf{maintenance time}
\item \textbf{middle-zone time} (etc.)
\end{enumerate}
Additionally, the Singular Value Decomposition of these empirical metrics is considered to define higher-level descriptors called \emph{styles}~\cite{chacoma2025data}.

\subsection{HodgeRank inference via metrics}

Given a metric, the coefficients
\[
f_{ij}
=
-
\frac{1}{2}(m_{ij}+m_{ji})
\]
are defined for each $i<j$.
The minus sign is imposed to identify ratings as potentials~\cite{perotti2025analysis}.
To test the HodgeRank inference method, two types of ground-truth coefficients are also introduced:  
(i) \emph{rank-based} coefficients $f_{ij}=i-j$, and  
(ii) \emph{points-based} coefficients $f_{ij}=\pi_j-\pi_i$.

In all cases, the underlying structure is a fully connected graph with $N$ nodes, promoted to a simplicial complex composed of $0$-simplices labeled by $i$, $1$-simplices labeled by $ij$ with $i<j$, and $2$-simplices labeled by $ijk$ with $i<j<k$.
Correspondingly, the coefficients $f_{ij}$ form the components of a $1$-cochain $f$, where the antisymmetry $f_{ji}=-f_{ij}$ is imposed by definition.
There are no $k$-simplices for $k>2$.
